A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-01-15 Manuscript/ohyou.tex
Addded figures to manuscript minus supplemental

One of the most commonly used Brownian-like models is the Ornstein Uhlenbeck (OU) model. This is a modification of the Brownian model with an additional parameter $\alpha$ that measures the strength of return towards a theoretical optimum \citep{hansen1997stabilizing}. The popularity of the OU model has grown exponentially in recent years (Fig. 1), in part because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}. OU models have become particularly prevalent in ecological studies: 652 papers containing the phrases “Ornstein Uhlenbeck” and “ecology” were published between 2011 and 2013 (Google Scholar search 30th January 2014). Additionally, although they are pattern based, fit to an OU model is now being used as evidence for the action of ecological, or ecologically driven, processes such as phylogenetic niche conservatism, convergent evolution and stabilizing selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}.

@@

  \item If $\alpha$ is small then evolution is approximately Brownian: If $\alpha$ is small then $1 - e^{-2\alpha T} \approx 2\alpha T$, i.e. traits accrue variance as if evolving according to a Brownian process. The question is obviously what is a small value of $\alpha$? Figure 2 provides some guidance on this issue, and see point (iv) below.\\

@@

Figure 2 gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure 2 highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. 2B).  Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. 2C). This will be indistinguishable from phylogenetically unstructured data.

In terms of statistical model fitting, a feature to highlight is that in Figure 2A and Figure 2C, changing the value of α across large relative ranges at the extremes (0 – 0.5 in Fig. 2A; 10-50 in Fig. 2C) produces relatively little quantitative change in the model behavior, and it is in the middle region of parameter space (1 – 5) that the biggest discernible effects occur.

@@

\subsection{Literature Review}
\label{section:literature.methods} 
To get an overview of the use of OU models in ecology and evolution, we used Google Scholar (accessed 21st November 2013) to locate papers published between 2005 \citep[when the R package ouch was released;][]{Butler:2004aa} and 2013 that contained the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” (the “biology” term was added to omit physics papers which also use the term “evolution”). We also recorded the total number of papers containing the terms “ecology”, or “evolution” and “biology” published between 2005 and 2013 and plotted the number of OU papers published each year as a proportion of the total number of papers published (Fig. 1).

@@

\section{Results}
\subsection{Simulations}
\label{section:sims.results} 
Figure 3 shows some examples of profile likelihoods for α generated on different datasets that were simulated according to a Brownian model. It is clear that the shape of the tree affects the shape of the likelihood profile. Trees with increasing speciation rates from root to tip (i.e., that have a lot of recently evolved species) generate very wide, flat likelihood profiles (Fig. 3A). On the other hand, trees with declining speciation rates from root to tip (Fig. 3B), or generated according to a Yule process (Fig. 3C), yield more strongly peaked profiles.

In all three cases, there are several features worth noting. First, the profiles are asymmetric. There is a hard bound at $\alpha = 0$, and as $\alpha$ becomes large, the rate of decrease in the likelihood slows. This is because, as shown in Figure 2, as values of $\alpha$ become large, the effects of changing $\alpha$ on model predictions are increasingly small. It is also notable, that the Maximum Likelihood estimates of $\alpha$ are, in most cases, between 0 and 1. In Figure 2 it was noted that in this region of parameter space, the OU model is very difficult to distinguish from Brownian. This, combined with the asymmetric nature of the profile likelihood is suggestive that estimates of $\alpha$ are likely to be biased, and that the asymptotic assumptions of likelihood ratio tests on $\alpha$ are likely to be violated.

The effects of the asymmetric profile likelihoods become clear when calculating Type I error for the OU model based on ML estimates across a range of tree shapes and sizes (Fig. 4; see Supplemental Figs. S2 and S3 for results with alternative priors for the Bayesian analyses). Regardless of tree shape, Type I error rates are unacceptably high when tree size is small (Figs. 4A, 4C and 4E). For some tree shapes (particularly where speciation rates accelerate towards the present), Type I error remains $>0.05$ even for trees with 1000 tips. Bayesian estimation is not subject to bias in the log-likelihood statistic and consistently rejects the OU model. However, estimates of α are similar regardless of method.

Figure 5 shows the proportion of data sets in which the OU model is favoured over the Brownian model for data simulated under Brownian motion with error (also see Figs. S4 and S5). The expectation is that the OU model should fit better because the branch length transformation partially captures the non-Brownian component (the error). There are two points worth noting. First, the frequency with which the OU model is favoured increases with tree size for both ML and Bayesian estimation (Fig. 5A). With as little as 5\% error, the OU model becomes extremely difficult to reject, even for trees with just 100 species. This is important for the interpretation of the OU model. We cannot conclude anything about evolutionary process from a single optimum OU model unless error is adequately accounted for. Second, for moderate amounts of error (5-10\%), the ML estimates of $\alpha$ are consistently $>$ 1 (Fig. 5B). Large values of $\alpha$ are similarly difficult to interpret because they are indicative that the signal of the past has been overwritten.

@@

\subsection{Literature Review}
\label{section:literature.results}
In total, 3720 papers containing the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” were published between 2005 and 2013, and the number has increased substantially since 2005 (Fig. 1). Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

@@

In terms of interpretation, it is worth pointing out that data will exhibit few signs of the limits to trait values that result from an OU process. This is because according to the OU process traits move back towards the optimum as their values become particularly large or small. Past history is effectively wiped over, as pointed out above, with the consequence that covariances between species become small (Fig. 2). Indeed, Figure 2 implies that the outcome of the OU process is indistinguishable from an acceleration of the rate of evolution towards the present. This interpretation is particular important for inference of phylogenetic niche conservatism. Although definitions of phylogenetic niche conservatism vary \citep{Losos:2008aa,Losos:2008ab,Wiens:2008aa,crisp2012phylogenetic} one proposed approach is to fit an OU model and take rejection of BM in favour of OU as evidence of constraints on niche evolution \citep{Wiens:2010aa}. A more parsimonious explanation would be that there is measurement error in the data. This seems particularly likely for climatic niche traits that are often averaged across an entire species range. Concluding from data on extant species that traits are limited by an OU-like process will therefore always be a leap of faith.

>>>>

One of the most commonly used Brownian-like models is the Ornstein Uhlenbeck (OU) model. This is a modification of the Brownian model with an additional parameter $\alpha$ that measures the strength of return towards a theoretical optimum \citep{hansen1997stabilizing}. The popularity of the OU model has grown exponentially in recent years (Fig. \ref{figure.literature}), in part because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}. OU models have become particularly prevalent in ecological studies: 652 papers containing the phrases “Ornstein Uhlenbeck” and “ecology” were published between 2011 and 2013 (Google Scholar search 30th January 2014). Additionally, although they are pattern based, fit to an OU model is now being used as evidence for the action of ecological, or ecologically driven, processes such as phylogenetic niche conservatism, convergent evolution and stabilizing selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}.

\begin{figure}
\centering
    \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure1.pdf}
\caption{The number of evolutionary biology (closed circles) and ecology (open circles) papers published between 2005 and 2012 containing the phrase “Ornstein-Uhlenbeck”, as a proportion of the total number of evolutionary biology or ecology papers published that year.}
\label{figure.literature}
\end{figure}

@@

  \item If $\alpha$ is small then evolution is approximately Brownian: If $\alpha$ is small then $1 - e^{-2\alpha T} \approx 2\alpha T$, i.e. traits accrue variance as if evolving according to a Brownian process. The question is obviously what is a small value of $\alpha$? Figure \ref{figure.traitsim} provides some guidance on this issue, and see point (iv) below.\\

@@

Figure \ref{figure.traitsim} gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure \ref{figure.traitsim} highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. \ref{figure.traitsim}B).  Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. \ref{figure.traitsim}C). This will be indistinguishable from phylogenetically unstructured data.

\begin{figure}
\centering
    \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure2.pdf}
\caption{Scaling of expected trait similarity with time since evolutionary divergence predicted by the Ornstein-Uhlenbeck model. The covariance between species’ traits values is scaled by the intra-specific trait variance (i.e. equal to correlation between species’ traits). This is plotted against the relative time of shared history (time at which species branched from each other, divided by the total tree height). }
\label{figure.traitsim}
\end{figure}

In terms of statistical model fitting, a feature to highlight is that in Figure \ref{figure.traitsim}A and Figure \ref{figure.traitsim}C, changing the value of α across large relative ranges at the extremes (0 – 0.5 in Fig. \ref{figure.traitsim}A; 10-50 in Fig. \ref{figure.traitsim}C) produces relatively little quantitative change in the model behavior, and it is in the middle region of parameter space (1 – 5) that the biggest discernible effects occur.

@@

\subsection{Literature Review}
\label{section:literature.methods} 
To get an overview of the use of OU models in ecology and evolution, we used Google Scholar (accessed 21st November 2013) to locate papers published between 2005 \citep[when the R package ouch was released;][]{Butler:2004aa} and 2013 that contained the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” (the “biology” term was added to omit physics papers which also use the term “evolution”). We also recorded the total number of papers containing the terms “ecology”, or “evolution” and “biology” published between 2005 and 2013 and plotted the number of OU papers published each year as a proportion of the total number of papers published (Fig. \ref{figure.literature}).

@@

\section{Results}
\subsection{Simulations}
\label{section:sims.results} 
Figure \ref{figure.profiles} shows some examples of profile likelihoods for α generated on different datasets that were simulated according to a Brownian model. It is clear that the shape of the tree affects the shape of the likelihood profile. Trees with increasing speciation rates from root to tip (i.e., that have a lot of recently evolved species) generate very wide, flat likelihood profiles (Fig. \ref{figure.profiles}A). On the other hand, trees with declining speciation rates from root to tip (Fig. \ref{figure.profiles}B), or generated according to a Yule process (Fig. \ref{figure.profiles}C), yield more strongly peaked profiles.

\begin{figure}
  \centering
  \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure3.pdf}
  \caption{Examples of profile likelihoods for selected simulated datasets. In all cases the ‘true’ value of alpha is 0.}
\label{figure.profiles}
\end{figure}

In all three cases, there are several features worth noting. First, the profiles are asymmetric. There is a hard bound at $\alpha = 0$, and as $\alpha$ becomes large, the rate of decrease in the likelihood slows. This is because, as shown in Figure \ref{figure.traitsim}, as values of $\alpha$ become large, the effects of changing $\alpha$ on model predictions are increasingly small. It is also notable, that the Maximum Likelihood estimates of $\alpha$ are, in most cases, between 0 and 1. In Figure \ref{figure.traitsim} it was noted that in this region of parameter space, the OU model is very difficult to distinguish from Brownian. This, combined with the asymmetric nature of the profile likelihood is suggestive that estimates of $\alpha$ are likely to be biased, and that the asymptotic assumptions of likelihood ratio tests on $\alpha$ are likely to be violated.

The effects of the asymmetric profile likelihoods become clear when calculating Type I error for the OU model based on ML estimates across a range of tree shapes and sizes (Fig. \ref{figure.rejection}; see Supplemental Figs. S2 and S3 for results with alternative priors for the Bayesian analyses). Regardless of tree shape, Type I error rates are unacceptably high when tree size is small (Figs. \ref{figure.rejection}A, \ref{figure.rejection}C and \ref{figure.rejection}E). For some tree shapes (particularly where speciation rates accelerate towards the present), Type I error remains $>0.05$ even for trees with 1000 tips. Bayesian estimation is not subject to bias in the log-likelihood statistic and consistently rejects the OU model. However, estimates of α are similar regardless of method.

\begin{figure}
  \centering
  \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure4.pdf}
  \caption{Rejection rates and estimation of $\alpha$ for different tree sizes and shapes. Rejection rates (a,c,e) for the OU model at multiple tree sizes for models fitted using ML (circles) and Bayesian (diamonds) methods. Estimates of $\alpha$ (b,d,f) based on ML (circles show means and dotted lines show 95\% sampling interval from 1000 trees) and Bayesian (diamonds show means and dotted lines show 95\% sampling interval of the modal value from the posterior distribution from 1000 trees).}
\label{figure.rejection}
\end{figure}

Figure \ref{figure.error} shows the proportion of data sets in which the OU model is favoured over the Brownian model for data simulated under Brownian motion with error (also see Figs. S4 and S5). The expectation is that the OU model should fit better because the branch length transformation partially captures the non-Brownian component (the error). There are two points worth noting. First, the frequency with which the OU model is favoured increases with tree size for both ML and Bayesian estimation (Fig. \ref{figure.error}A). With as little as 5\% error, the OU model becomes extremely difficult to reject, even for trees with just 100 species. This is important for the interpretation of the OU model. We cannot conclude anything about evolutionary process from a single optimum OU model unless error is adequately accounted for. Second, for moderate amounts of error (5-10\%), the ML estimates of $\alpha$ are consistently $>$ 1 (Fig. \ref{figure.error}B). Large values of $\alpha$ are similarly difficult to interpret because they are indicative that the signal of the past has been overwritten.

\begin{figure}
  \centering
  \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure5.pdf}
  \caption{Rejection of BM with measurement error (a) and estimation of α (b). Symbols follow Figure \ref{figure.rejection}.}
\label{figure.error}
\end{figure}

@@

\subsection{Literature Review}
\label{section:literature.results}
In total, 3720 papers containing the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” were published between 2005 and 2013, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

@@

In terms of interpretation, it is worth pointing out that data will exhibit few signs of the limits to trait values that result from an OU process. This is because according to the OU process traits move back towards the optimum as their values become particularly large or small. Past history is effectively wiped over, as pointed out above, with the consequence that covariances between species become small (Fig. \ref{figure.traitsim}). Indeed, Figure \ref{figure.traitsim} implies that the outcome of the OU process is indistinguishable from an acceleration of the rate of evolution towards the present. This interpretation is particular important for inference of phylogenetic niche conservatism. Although definitions of phylogenetic niche conservatism vary \citep{Losos:2008aa,Losos:2008ab,Wiens:2008aa,crisp2012phylogenetic} one proposed approach is to fit an OU model and take rejection of BM in favour of OU as evidence of constraints on niche evolution \citep{Wiens:2010aa}. A more parsimonious explanation would be that there is measurement error in the data. This seems particularly likely for climatic niche traits that are often averaged across an entire species range. Concluding from data on extant species that traits are limited by an OU-like process will therefore always be a leap of faith.
